A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-09 podc_paper/algorithm-2.tex
got the bound on the size of the RBTs

The space-bounded implementation uses an additional shared array so that processes can access one another's leaves to perform helping.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\end{itemize}
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}
Each \typ{Node} has an additional array field to help determine which of its blocks are safe to discard.
\begin{itemize}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the node that process $k$ }\\
\mbox{ }\Comment{has accessed, either to find the enqueued item for its dequeue}\\
\mbox{ }\Comment{or during a \nl\ dequeue}
\end{itemize}
The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
Each RBT is initialized with an empty block with index 0.
The \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
longer needed; we can instead find the maximum \fld{index} of any block in the RBT.
To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
returns the block with the maximum \fld{index}.
We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
can be done in constant time, without affecting the time of other RBT operations.

@@

\HERE{Update this parag after making changes}
Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode3GC}.
The \op{Propagate} and \op{FindResponse} routines are unchanged.
Minor additions to \op{IndexDequeue} and \op{GetEnqueue} update the \var{last} field of nodes
visited by the routines to mark the blocks they visit in those nodes.
Only minor modifications have been made to \op{Enqueue}, \op{Append}, \op{CreateBlock}
to accomodate the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
of the \head\ field).
\op{Dequeue} has been similarly modified, and the second half of the routine is now in a
separate routine called \op{CompleteDeq} so that it can also be used by other processes
helping to complete the operation.

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{250}
\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $G$}
    	\State \linecomment{Do garbage collection}\label{GCstart}
        \State $s \assign \Call{SplitBlock}{v}.\var{index}$
        \State \Call{Help}{}
        \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
        \State \Return \Call{Insert}{$T',B$}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}
    \EndIf
\EndFunction{AddBlock}
}
\spac

@@

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{353}
\gc{
\Function{Block}{SplitBlock}{\typ{Node} v}
	\State \linecomment{figure out where to split $v$'s RBT}
	\If{$v=\var{root}$}
		\State $m \assign 0$
		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
		\EndFor
		\State $B\assign \var{root}.\fld{blocks}[m-1]$
	\Else
		\State $B_p \assign \Call{SplitBlock}{v.parent}$
		\State \var{dir} \assign ($v=v.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B \assign v.\fld{blocks}[B_p.\edir]$
	\EndIf
	\State \Return ($B = \nl$ ? \Call{MinBlock}{v.\var{blocks}} : $B$)
\EndFunction{SplitBlock}
}

>>>>

The space-bounded implementation uses two additional shared arrays:
the \var{leaf} array allows processes to access one another's leaves to perform helping, and
the \var{last} array is used to determine which blocks are safe to discard.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ is the largest \fld{index} of a block in the root  that process $k$}\\
\mbox{ }\Comment{has observed to contain either an enqueue whose value has}\\
\mbox{ }\Comment{been dequeued or a null dequeue }
\end{itemize}
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}
The \fld{blocks} field of each node in the \ordering\ tree is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.  
Each RBT is initialized with an empty block with index 0.
The \head\ field, which previously gave the next position to insert into the \fld{blocks} array is no
longer needed; we can instead find the maximum \fld{index} of any block in the RBT.
To facilitate this, \op{MaxBlock} is a query operation on the RBT that 
returns the block with the maximum \fld{index}.
We can store, in the root of the RBT, a pointer to the maximum block so that \op{MaxBlock}
can be done in constant time, without affecting the time of other RBT operations.
Similarly, a \op{MinBlock} query finds the block with the minimum \fld{index} in a RBT.

@@

\here{Update this parag after code is finalized--it has changed since I wrote this parag}
Pseudocode for the space-bounded implementation appears in Figure \ref{pseudocode1GC}--\ref{pseudocode2GC}.
The \op{Propagate} and \op{FindResponse} routines are unchanged.
Minor additions to \op{IndexDequeue} and \op{GetEnqueue} update the \var{last} field of nodes
visited by the routines to mark the blocks they visit in those nodes.
Only minor modifications have been made to \op{Enqueue}, \op{Append}, \op{CreateBlock}
to accomodate the switch from an array of blocks to a RBT of blocks (and the corresponding disappearance
of the \head\ field).
\op{Dequeue} has been similarly modified, and the second half of the routine is now in a
separate routine called \op{CompleteDeq} so that it can also be used by other processes
helping to complete the operation.

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{250}
\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $G$}
    	\State \linecomment{Do garbage collection}\label{GCstart}
        \State $s \assign \Call{SplitBlock}{v}.\var{index}$\label{callSplitBlock}
        \State \Call{Help}{}
        \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
        \State \linecomment{\op{Split} removes blocks with $\fld{index} < m-1$}
        \State \Return \Call{Insert}{$T',B$}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}
    \EndIf
\EndFunction{AddBlock}
}
\spac

@@

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{353}
\gc{
\Function{Block}{SplitBlock}{\typ{Node} v}
	\State \linecomment{figure out where to split $v$'s RBT}
	\If{$v=\var{root}$}
		\State $m \assign 0$
		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
		\EndFor
		\State $B\assign \var{root}.\fld{blocks}[m-1]$
	\Else
		\State $B_p \assign \Call{SplitBlock}{v.parent}$
		\State \var{dir} \assign ($v=v.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B \assign v.\fld{blocks}[B_p.\edir]$
	\EndIf
	\State \linecomment If $B$ was already discarded, use leftmost block instead
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{v.\var{blocks}}$ : $B$)
\EndFunction{SplitBlock}
}

### 2023-01-09 podc_paper/correctness-2.tex
got the bound on the size of the RBTs

Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in $v$'s tree at some time during the execution.
It follows from \Cref{RBTupdates} and \Cref{nonemptyConsec} that 
each time a new block appears in $v$'s RBT, its \fld{index} is greater than any block
that has appeared in $v$'s RBT earlier.
Thus, $v.\blocks[b]$ is unique, if it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.
We describe later what should happen if that block is not found by the search because it has been removed
by a GC phase.
\here{remember to come back to this point}

@@

\here{next, should probably show that blocks are removed only after they are no longer needed; i.e., every dequeue has its answer (either returned, or written in its response field) and every enqueue has a matching dequeue that has either terminated or has a response written.}

-------

\subsection{Analysis}

We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
Recall that GC is done on a node every $G$ times its RBT is updated, where we chose $G$ to be $p^2\ceil{\log p}$.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains $O(q_{max}+p)$ blocks.
\end{lemma}
\begin{proof}
\here{to be written}
\end{proof}

@@

\begin{theorem}\label{spaceBound}
The size of our queue data structure is $O(pq_{max}+p^3\log p)$ words of memory.
\end{theorem}
\begin{proof}
There are $2p-1$ nodes in the ordering tree.  Aside from the RBT, each node uses $O(p)$ memory for the
\fld{last} array.  So the space bound follows from \Cref{RBTbound} and the fact that $G$ is chosen
to be $p^2\ceil{\log p}$.
\end{proof}

@@

\begin{proposition}
The amortized step complexity is $O(\log p \log(p+q_{max}))$ per operation.
\end{proposition}
\begin{proof}
It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can all be done in $O(\log(p+q_{max}))$ steps.

>>>>

Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in $v$'s tree at some time during the execution.
It follows from \Cref{RBTupdates} and \Cref{nonEmptyIncreasing} that 
each time a new block appears in $v$'s RBT, its \fld{index} is greater than any block
that has appeared in $v$'s RBT earlier.
Thus, $v.\blocks[b]$ is unique, if it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.
We describe later what should happen if that block is not found by the search because it has been removed
by a GC phase.
\here{remember to come back to this point}

@@

\here{next, should probably show that blocks are removed only after they are no longer needed; i.e., every dequeue has its answer (either returned, or written in its response field) and every enqueue has a matching dequeue that has either terminated or has a response written.}
one invariant will be that a null dequeue or a dequeued enqueue appears in the first block (or some later block) of each RBT.
should imply that no operation will need any block before the first block in the RBT
-------

\subsection{Analysis}
\label{sec::GCanalysis}

\here{need to prove a lemma analogous to \Cref{blockSize}}

We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.

%We say the index $b$ \emph{is marked in node $v$} if some process has tested
%whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
%Intuitively, a block's index value is marked 
%when a dequeue passes through this block to find the enqueue
%whose argument it should return, or when a dequeue that returns null passes through the block.
%Thus, $b$ gets marked in $v$ when $v.\fld{blocks}[b]$ 
%contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
%If the index of block $B$ is marked in $v$, we also say $B$ is marked in $v$.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $2q_{max}+4p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node $v$'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
$A$ performs a GC phase.
Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $C$.

We first bound the number of operations in $L_2$.

The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
Any enqueue in $L_2$ whose value is not still in the queue at the end of $L_2$
must still be in progress at $C$; otherwise the process that dequeued it would have
set its \var{last} entry to a value greater than the index of the block of the root that contains the enqueue prior to $C$,
contradicting the definition of $m$.
So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
Thus, there are at most $q_{max}+p$ enqueues in $L_2$.

If a dequeue in $L_2$ returns a non-null value in the sequential execution $L$,
then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
Thus, there are at most $q_{max}+(q_{max}+p)$ dequeues in $L_2$ that return non-null values.
Any dequeue in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
at $C$; otherwise the process that performed the dequeue would have set its \var{last} entry
to a value greater than the index of the block of the root that contains the dequeue prior to $C$, 
contradicting the definition of $m$.
So, there are at most $p$ null dequeues in $L_2$.
Thus, there are at most $2q_{max} + 2p$ dequeues in $L_2$.

$A$'s call to \op{SplitBlock} determines the index $s$ used to split $v$'s RBT by following 
\eleft\ and \eright\ pointers from the root down to $v$.
So, the block returned is a subblock of block $m$ of the root, unless
at some point along the path of subblocks the subblock has already been removed by a split,
in which case \op{SplitBlock} returns a subblock of a block $m'>m$ in the root.

Next, we bound the number of operations in $v$'s blocks that are retained when $A$ does sets
$T'\assign\op{Split}(T,s)$ operation.
Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$.\here{cite a lemma about propagation before operation terminates}
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
There are at most $p$ operations in block $m$ of the root by \Cref{blockSizeGC}
and we showed above that there are at most $2q_{max} + 2p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $2q_{max}+4p$ operations in blocks of $T'$.
Since each block is non-empty by \Cref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
block is inserted before $A$ sets $v$'s \fld{blocks} to the resulting RBT.
\end{proof}

@@

\begin{theorem}\label{spaceBound}
The size of our queue data structure is $O(pq_{max}+p^3\log p)$ words of memory.
\end{theorem}
\begin{proof}
There are $2p-1$ nodes in the ordering tree.  Aside from the RBT, each node uses $O(1)$ memory words.
So the space bound follows from \Cref{RBTbound} and the fact that $G$ is chosen
to be $p^2\ceil{\log p}$.
\end{proof}

@@

\begin{proposition}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{proposition}
\begin{proof}
\here{this proof may need some modification since code was changed to change how split point is computed}
It follows from \Cref{RBTbound} and the fact that $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can all be done in $O(\log(p+q_{max}))$ steps.
